---
title: "How pclmbayes compares with the ungroup package"
author: "pclmbayes authors"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{How pclmbayes compares with the ungroup package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center"
)
set.seed(2026)
have_ungroup <- requireNamespace("ungroup", quietly = TRUE)
```

# Why a separate vignette?

`pclmbayes` and the popular `ungroup` package both implement
*penalised composite link models* (PCLM) for redistributing
wide-bin counts onto a fine grid. They share the high-level idea —
B-splines plus a difference penalty plus a composite-link `C` matrix —
but they implement **different statistical models** with different
likelihoods, links, identifiability conventions, and supported bin
layouts. This vignette spells the differences out, then compares
all three algorithms shipped with `pclmbayes` (`pclm()`, `bpclm()`,
`pclm_exact()`) against `ungroup::pclm()` on synthetic data with
known truth and on the two real datasets distributed with the
package.

```{r}
library(pclmbayes)
```

# 1. Algorithmic differences (theory)

| Aspect | `pclmbayes` (`pclm`, `bpclm`, `pclm_exact`) | `ungroup::pclm` |
|---|---|---|
| Reference | Lambert & Eilers (2009) CSDA | Eilers (2007) / Rizzi et al. (2015) |
| Likelihood | **Multinomial** on wide-bin counts $m_j$ | **Poisson** on wide-bin counts $y_j$ |
| Link | **Softmax**: $\pi_i = e^{(B\phi)_i} / \sum_\ell e^{(B\phi)_\ell}$ | **Log**: $\mu_i = e^{(B\phi)_i}$ |
| What is estimated | Probability density $\pi_i$ that integrates to 1 | Non-negative count intensity $\mu_i$ on the fine grid |
| Identifiability | Constraint $\sum_k \phi_k = 0$ enforced each iteration (softmax is shift-invariant) | None needed (intercept absorbed by the log link) |
| Smoothing parameter | Grid search over BIC (default) or AIC (`pclm`); Gibbs draw for `bpclm`; not needed for `pclm_exact` | Numerical optimisation (`stats::optimise`) of AIC or BIC |
| Bin layout | Arbitrary, possibly **overlapping** wide bins via the rectangle method (`bin_matrix()`) | Strictly **contiguous** non-overlapping bins, output grid commensurable with bin widths |
| Exposure / offset | Not built-in (model is for the density) | Optional `offset` for population-at-risk |
| Exact preservation of bin totals | `pclm_exact()` (constrained MAP) and `calibrate()` | Not exposed as an option (it is a *smoothed* fit) |
| Shape constraints (unimodal, log-concave, monotonic) | Yes, in `bpclm()` (Eq. 7 of the paper) | No |
| Full posterior inference | Yes (`bpclm`, MALA + Gibbs) | No (frequentist GLM with delta-method CI) |
| Native dimension | 1D | 1D and 2D (e.g. age × year mortality surfaces) |

In short: `ungroup` is a Poisson smoother for grouped counts, while
`pclmbayes` is a multinomial density estimator with a Bayesian variant
and an exact-ungrouping option. The two solve overlapping but distinct
problems; numerical results agree closely for the *shape* of the fitted
distribution, but disagree by O(shrinkage) on the bin totals.

# 2. The three algorithms inside `pclmbayes`

Before contrasting with `ungroup`, recap what each `pclmbayes` entry
point does (all use the same multinomial / softmax model):

* **`pclm()`** — frequentist penalised scoring (Newton/IRLS) that
  maximises the penalised multinomial log-likelihood. The smoothing
  parameter $\tau$ is selected by BIC over a default grid. Returns a
  point estimate $\hat\phi$ and its asymptotic
  variance-covariance matrix from the inverse penalised Fisher
  information.
* **`bpclm()`** — fully Bayesian: MALA proposal for $\phi$ with adaptive
  step size targeting 0.57 acceptance, Gibbs draw for $\tau$ from a
  Gamma full conditional. Returns a chain of $\phi$, $\pi$ and $\tau$.
  Optionally enforces unimodality, log-concavity, or monotonicity by
  rejecting proposals that violate them (Eq. 7 of Lambert & Eilers).
* **`pclm_exact()`** — constrained MAP. Solves
  $\min_\phi \tfrac12\,\phi'P\phi$
  subject to $C\pi(\phi) = m / m_+$ exactly, by sequential quadratic
  programming on the Lagrangian. The smoothness penalty is the *only*
  objective; the bin counts are reproduced to machine precision. This
  is the canonical "ungrouping" point estimate.

The three differ in *what they trade off*. `pclm()` and `bpclm()`
trade fit-to-bins against smoothness through $\tau$; `pclm_exact()`
hits the bins exactly and minimises the residual roughness. `bpclm()`
adds posterior uncertainty.

# 3. Synthetic comparison: 5-year mortality bands → single years

This is the canonical demographic ungrouping use case: we know the
truth (single-year deaths), aggregate to 5-year bands, and recover the
single-year breakdown. We measure RMSE per single-year cell against
the truth, max absolute residual on the wide-bin totals, and runtime.

```{r synth-data}
ages   <- 0:99
true_p <- 0.10 * dnorm(ages, mean =  2, sd =  3) +
          0.06 * dnorm(ages, mean = 25, sd = 12) +
          0.84 * dnorm(ages, mean = 75, sd = 12)
true_p <- true_p / sum(true_p)
N      <- 50000L
true_yearly <- as.numeric(rmultinom(1L, size = N, prob = true_p))
year_to_band <- (ages %/% 5L) + 1L
m_band <- as.numeric(tapply(true_yearly, year_to_band, sum))
wb     <- cbind(seq(0, 95, by = 5L), seq(5, 100, by = 5L))
```

## Fits

```{r synth-fits}
t_pclm <- system.time(
  fit_pclm <- pclm(m = m_band, wide_breaks = wb,
                   a = 0, b = 100, ngrid = 100L,
                   ndx = 22L, degree = 3L, penalty_order = 3L)
)[["elapsed"]]

t_pclm_exact <- system.time(
  fit_exact <- pclm_exact(m = m_band, wide_breaks = wb,
                          a = 0, b = 100, ngrid = 100L,
                          ndx = 22L, degree = 3L, penalty_order = 3L)
)[["elapsed"]]

t_bpclm <- system.time(
  fit_b <- bpclm(m = m_band, wide_breaks = wb,
                 a = 0, b = 100, ngrid = 100L,
                 ndx = 22L, degree = 3L, penalty_order = 3L,
                 niter = 3000L, burnin = 1000L, adapt = 500L,
                 seed = 7L)
)[["elapsed"]]
fit_b_cal <- calibrate(fit_b)             # exact band totals
```

```{r synth-ungroup, eval = have_ungroup}
t_ungroup <- system.time(
  fit_ung <- ungroup::pclm(x = wb[, 1L], y = m_band,
                           nlast = 5, out.step = 1)
)[["elapsed"]]
ung_yearly <- as.numeric(fit_ung$fitted)  # length 100, single-year estimates
```

```{r synth-ungroup-skip, eval = !have_ungroup, echo = FALSE}
message("ungroup not installed; skipping ungroup::pclm() comparison.")
t_ungroup  <- NA_real_
ung_yearly <- rep(NA_real_, length(ages))
```

## Per-cell RMSE and bin-total residuals

```{r synth-metrics}
yhat_pclm  <- fit_pclm$pi  * sum(m_band)
yhat_exact <- fit_exact$pi * sum(m_band)
yhat_bpclm <- fit_b_cal$pi * sum(m_band)

rmse <- function(a, b) sqrt(mean((a - b) ^ 2))

# Wide-bin residual after re-aggregating the single-year fit
band_resid <- function(yhat) {
  band <- as.numeric(tapply(yhat, year_to_band, sum))
  max(abs(band - m_band))
}

results <- data.frame(
  algorithm   = c("pclm",          "bpclm + calibrate", "pclm_exact",
                  "ungroup::pclm"),
  RMSE_yearly = c(rmse(yhat_pclm,  true_yearly),
                  rmse(yhat_bpclm, true_yearly),
                  rmse(yhat_exact, true_yearly),
                  rmse(ung_yearly, true_yearly)),
  max_bin_residual = c(band_resid(yhat_pclm),
                       band_resid(yhat_bpclm),
                       band_resid(yhat_exact),
                       band_resid(ung_yearly)),
  elapsed_sec = c(t_pclm, t_bpclm, t_pclm_exact, t_ungroup)
)
results
```

The expected pattern:

* `pclm()` — the multinomial smoother — has a **non-zero** max bin
  residual (the smoothness penalty pulls bin probabilities away from
  the empirical proportions). Its per-cell RMSE is competitive but it
  is the *wrong tool* if you want exact totals.
* `pclm_exact()` and `bpclm() + calibrate()` reproduce bin totals to
  machine precision. They are the right tool for ungrouping.
* `ungroup::pclm()` — the Poisson smoother — also tends to have a
  small but non-zero bin residual (the Poisson PCLM is a smoothed fit,
  not a constrained one).
* RMSE values across all four are typically within 5–10% of each other
  on this size of data; the per-cell shape is dominated by the
  irreducible multinomial-within-band sampling noise of order
  $\sqrt{n_j / 6}$ for a 5-year band.

## Plot the four point estimates

```{r synth-plot, fig.cap = "Single-year fits from the four algorithms vs. the simulated truth."}
plot(ages, true_yearly, type = "h", col = "grey70",
     xlab = "Age (years)", ylab = "Deaths",
     main = "Recovered single-year counts")
lines(ages, yhat_exact, col = "black",     lwd = 2)
lines(ages, yhat_pclm,  col = "steelblue", lwd = 2, lty = 2)
lines(ages, yhat_bpclm, col = "firebrick", lwd = 2, lty = 3)
if (have_ungroup) {
  lines(ages, ung_yearly, col = "darkgreen", lwd = 2, lty = 4)
  legend("topleft", bty = "n",
         legend = c("Truth", "pclm_exact", "pclm",
                    "bpclm + calibrate", "ungroup::pclm"),
         col    = c("grey70", "black", "steelblue",
                    "firebrick", "darkgreen"),
         lwd    = c(1, 2, 2, 2, 2),
         lty    = c(1, 1, 2, 3, 4))
} else {
  legend("topleft", bty = "n",
         legend = c("Truth", "pclm_exact", "pclm", "bpclm + calibrate"),
         col    = c("grey70", "black", "steelblue", "firebrick"),
         lwd    = c(1, 2, 2, 2),
         lty    = c(1, 1, 2, 3))
}
```

Visually all four are nearly indistinguishable in the body of the
distribution; the perceptible differences are at the boundaries (the
Poisson and softmax tails fall off slightly differently) and inside
bins where `pclm_exact()` lacks the small kinks that `calibrate()` can
introduce.

## Uncertainty (only `bpclm` provides it natively)

`pclm()`, `pclm_exact()` and `ungroup::pclm()` are point estimates.
Of the four, only `bpclm()` produces a *posterior chain* of densities
that, after `calibrate()`, all preserve the bin totals exactly — so
posterior predictive intervals at single-year resolution come for
free. `ungroup::pclm()` returns delta-method confidence intervals
based on the asymptotic variance of the Poisson PCLM, but has no
constraint that the resulting interval respect the band totals.

```{r synth-pp, fig.cap = "Posterior predictive 90% PI from bpclm + calibrate."}
pp <- posterior_predict(fit_b_cal, type = "predictive",
                        level = 0.9, seed = 7L)
plot(pp, show_bins = FALSE,
     xlab = "Age (years)", ylab = "Single-year deaths",
     main = "bpclm + calibrate: 90% posterior PI")
lines(ages, true_yearly, col = "grey50", lwd = 1.4, lty = 2)
```

# 4. Real data — bloodlead (Hasselblad et al., 1980)

Seven irregular wide bins of blood-lead concentration, $n = 139$. No
ground truth — we visualise the recovered density.

```{r blood-fits}
data(bloodlead)
wb_b <- with(bloodlead, cbind(lower, upper))

fit_b_pclm  <- pclm(      m = bloodlead$count, wide_breaks = wb_b,
                          a = 0, b = 80, ngrid = 80L, ndx = 17L,
                          degree = 3L, penalty_order = 3L)
fit_b_exact <- pclm_exact(m = bloodlead$count, wide_breaks = wb_b,
                          a = 0, b = 80, ngrid = 80L, ndx = 17L,
                          degree = 3L, penalty_order = 3L)
fit_b_bayes <- bpclm(     m = bloodlead$count, wide_breaks = wb_b,
                          a = 0, b = 80, ngrid = 80L, ndx = 17L,
                          degree = 3L, penalty_order = 3L,
                          niter = 4000L, burnin = 1000L, adapt = 500L,
                          shape = "unimodal", seed = 1L)
```

```{r blood-ungroup, eval = have_ungroup}
# bloodlead bins are contiguous: x = lower limits, nlast = last bin width.
fit_b_ung <- ungroup::pclm(x        = bloodlead$lower,
                           y        = bloodlead$count,
                           nlast    = 80 - 65,   # last bin is [65, 80)
                           out.step = 1)
ung_density_b <- as.numeric(fit_b_ung$fitted) / sum(fit_b_ung$fitted)
```

```{r blood-plot, fig.cap = "Recovered density on the bloodlead data, from all four algorithms."}
mids_b <- (head(fit_b_pclm$grid, -1L) + tail(fit_b_pclm$grid, -1L)) / 2
plot(mids_b, fit_b_pclm$pi,  type = "l", lwd = 2, lty = 2,
     col = "steelblue",
     xlab = "Blood-lead concentration (µg/dl)",
     ylab = "Density mass per fine cell",
     main = "bloodlead: density estimates")
lines(mids_b, fit_b_exact$pi, col = "black",     lwd = 2)
lines(mids_b, fit_b_bayes$pi, col = "firebrick", lwd = 2, lty = 3)
if (have_ungroup) {
  lines(seq(0.5, 79.5, by = 1), ung_density_b,
        col = "darkgreen", lwd = 2, lty = 4)
  legend("topright", bty = "n",
         legend = c("pclm_exact", "pclm", "bpclm (unimodal)",
                    "ungroup::pclm"),
         col = c("black", "steelblue", "firebrick", "darkgreen"),
         lwd = 2, lty = c(1, 2, 3, 4))
} else {
  legend("topright", bty = "n",
         legend = c("pclm_exact", "pclm", "bpclm (unimodal)"),
         col = c("black", "steelblue", "firebrick"),
         lwd = 2, lty = c(1, 2, 3))
}
```

```{r blood-bins}
# How well does each fit reproduce the seven observed bin totals?
band_tot <- function(pi_grid, fit_obj) {
  as.numeric(fit_obj$C %*% pi_grid) * sum(fit_obj$m)
}
bl_compare <- data.frame(
  observed       = bloodlead$count,
  pclm           = round(fit_b_pclm$fitted_counts,  2),
  bpclm          = round(band_tot(fit_b_bayes$pi,   fit_b_bayes),  2),
  pclm_exact     = round(fit_b_exact$fitted_counts, 6),
  ungroup        = if (have_ungroup) {
                     # Re-aggregate ungroup's single-year output back to bins
                     yhat <- as.numeric(fit_b_ung$fitted)
                     bin_id <- findInterval(seq(0.5, 79.5, by = 1),
                                            c(bloodlead$lower, 80),
                                            rightmost.closed = TRUE)
                     round(as.numeric(tapply(yhat, bin_id, sum)), 2)
                   } else {
                     rep(NA_real_, nrow(bloodlead))
                   }
)
bl_compare
```

`pclm_exact` reproduces the seven observed counts to machine
precision; `pclm()` and `bpclm()` shrink the largest bin by ~10%
toward the smooth fit; `ungroup::pclm()` is also a smoothed fit and
has a similar (but generally smaller, owing to the Poisson rather
than multinomial likelihood) shrinkage pattern.

# 5. Real data — tbdeaths1907 (illustrative)

Twelve irregular age bands of tuberculosis deaths in The Netherlands,
1907 (illustrative reconstruction; see `?tbdeaths1907`). Many more
counts ($n = 9440$) than `bloodlead`, so smoothing-induced shrinkage
is small in absolute terms.

```{r tb-fits}
data(tbdeaths1907)
wb_t <- with(tbdeaths1907, cbind(lower, upper))

fit_t_pclm  <- pclm(      m = tbdeaths1907$count, wide_breaks = wb_t,
                          a = 0, b = 120, ngrid = 120L, ndx = 17L,
                          degree = 3L, penalty_order = 3L)
fit_t_exact <- pclm_exact(m = tbdeaths1907$count, wide_breaks = wb_t,
                          a = 0, b = 120, ngrid = 120L, ndx = 17L,
                          degree = 3L, penalty_order = 3L)
fit_t_bayes <- bpclm(     m = tbdeaths1907$count, wide_breaks = wb_t,
                          a = 0, b = 120, ngrid = 120L, ndx = 17L,
                          degree = 3L, penalty_order = 3L,
                          niter = 3000L, burnin = 1000L, adapt = 500L,
                          seed = 2L)
```

```{r tb-ungroup, eval = have_ungroup}
# tbdeaths1907 bins are contiguous; last bin is [100, 120).
fit_t_ung <- ungroup::pclm(x        = tbdeaths1907$lower,
                           y        = tbdeaths1907$count,
                           nlast    = 120 - 100,
                           out.step = 1)
ung_density_t <- as.numeric(fit_t_ung$fitted) / sum(fit_t_ung$fitted)
```

```{r tb-plot, fig.cap = "Recovered age-at-death density on tbdeaths1907."}
mids_t <- (head(fit_t_pclm$grid, -1L) + tail(fit_t_pclm$grid, -1L)) / 2
plot(mids_t, fit_t_pclm$pi,  type = "l", lwd = 2, lty = 2,
     col = "steelblue",
     xlab = "Age (years)",
     ylab = "Density mass per single year",
     main = "tbdeaths1907: density estimates")
lines(mids_t, fit_t_exact$pi, col = "black",     lwd = 2)
lines(mids_t, fit_t_bayes$pi, col = "firebrick", lwd = 2, lty = 3)
if (have_ungroup) {
  lines(seq(0.5, 119.5, by = 1), ung_density_t,
        col = "darkgreen", lwd = 2, lty = 4)
  legend("topright", bty = "n",
         legend = c("pclm_exact", "pclm", "bpclm",
                    "ungroup::pclm"),
         col = c("black", "steelblue", "firebrick", "darkgreen"),
         lwd = 2, lty = c(1, 2, 3, 4))
} else {
  legend("topright", bty = "n",
         legend = c("pclm_exact", "pclm", "bpclm"),
         col = c("black", "steelblue", "firebrick"),
         lwd = 2, lty = c(1, 2, 3))
}
```

# 6. Summary — when to use what

* **You want a smooth density and don't care about preserving bin
  totals exactly** — either `pclm()` (multinomial) or
  `ungroup::pclm()` (Poisson) will do; pick the one whose model
  matches your data-generating story. They give very similar shapes.
* **You want exact preservation of wide-bin totals** (the canonical
  ungrouping use case) — use `pclm_exact()`. Neither `ungroup::pclm()`
  nor `pclm()` does this.
* **You want full posterior uncertainty, optionally with a shape
  constraint (unimodal, log-concave, monotonic)** — use `bpclm()`,
  optionally followed by `calibrate()` to enforce exact band totals on
  every draw. `ungroup` does not provide this.
* **You have overlapping or misaligned wide bins** — only
  `pclmbayes` supports this (via the rectangle method in
  `bin_matrix()`). `ungroup::pclm()` requires contiguous bins.
* **You have a 2D problem (e.g. age × year mortality)** — only
  `ungroup` provides it directly (`ungroup::pclm2D()`). `pclmbayes` is
  1D in the current version.
* **You need an exposure / population-at-risk offset** — `ungroup`
  has it built in; in `pclmbayes` you would need to model rates
  externally.

# References

Eilers, P. H. C. (2007). Ill-posed problems with counts, the
composite link model and penalized likelihood. *Statistical
Modelling*, 7(3), 239–254.

Eilers, P. H. C. and Marx, B. D. (1996). Flexible smoothing with
B-splines and penalties. *Statistical Science*, 11(2), 89–121.

Hasselblad, V., Stead, A. G. and Galke, W. (1980). Analysis of
coarsely grouped data from the lognormal distribution. *Journal of
the American Statistical Association*, 75, 771–778.

Lambert, P. and Eilers, P. H. C. (2009). Bayesian density estimation
from grouped continuous data. *Computational Statistics and Data
Analysis*, 53(4), 1388–1399. doi:10.1016/j.csda.2008.11.022

Pascariu, M. D., Dańko, M. J., Schöley, J. and Rizzi, S. (2018).
ungroup: An R package for efficient estimation of smooth distributions
from coarsely binned data. *Journal of Open Source Software*, 3(29),
937.

Rizzi, S., Thinggaard, M., Engholm, G., Christensen, N. C., Jacobsen,
R., Vaupel, J. W. and Lindahl-Jacobsen, R. (2015). Efficient estimation
of smooth distributions from coarsely grouped data. *American Journal
of Epidemiology*, 182(2), 138–147.
