---
title: "Bayesian density estimation and exact ungrouping with pclmbayes"
author: "pclmbayes authors"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Bayesian density estimation and exact ungrouping with pclmbayes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.align = "center"
)
set.seed(2026)
```

# Overview

The **pclmbayes** package estimates a smooth latent continuous density
from data that have only been observed in wide bins. Two paradigms are
supported and they answer two different questions.

1. **Density estimation** (the original paradigm of Lambert and Eilers,
   2009). The wide-bin counts are treated as a multinomial *sample* of
   an underlying smooth density, which is recovered with the
   composite-link spline model. The fitted bin counts are *not*
   constrained to match the observed bin counts exactly — the
   smoothness penalty is allowed to shrink them.
2. **Exact ungrouping** (the use case for distributing wide-bin counts
   across a fine grid — for example, redistributing deaths in 5-year
   age bands across single years of age). The wide-bin counts are
   treated as *fixed totals* and the fine-grid counts must sum back to
   them exactly.

The package provides functions for both. The decision tree is:

* `pclm()` — frequentist density estimation (BIC-selected smoothing).
* `bpclm()` — Bayesian density estimation with shape constraints
  (unimodal, log-concave, monotonic) and full posterior uncertainty.
* `pclm_exact()` — frequentist constrained MAP that gives the smoothest
  density consistent with the observed bin totals exactly.
* `calibrate()` — post-hoc projection of any fit (frequentist or
  Bayesian) onto the constraint manifold (exact bin totals).
* `posterior_predict()` — credible/prediction intervals for fine-cell
  counts, with multinomial sampling noise included where appropriate.

```{r}
library(pclmbayes)
```

# Part 1 — Density estimation (Lambert & Eilers paradigm)

## The bloodlead example (paper, §6.1)

Seven wide bins of blood lead concentration in young Puerto Rican
children in 1974 (Hasselblad et al. 1980, reproduced verbatim in
Lambert and Eilers 2009).

```{r}
data(bloodlead)
bloodlead
```

### Frequentist fit

```{r}
fit_f <- pclm(
  m             = bloodlead$count,
  wide_breaks   = with(bloodlead, cbind(lower, upper)),
  a             = 0, b = 80,
  ngrid         = 80L,
  ndx           = 17L, degree = 3L,
  penalty_order = 3L,
  select        = "BIC"
)
fit_f
```

```{r, fig.cap = "Frequentist (BIC-smoothed) PCLM fit to the bloodlead data."}
plot(fit_f)
```

### Bayesian fit with unimodality constraint

```{r}
fit_b <- bpclm(
  m             = bloodlead$count,
  wide_breaks   = with(bloodlead, cbind(lower, upper)),
  a             = 0, b = 80,
  ngrid         = 80L,
  ndx           = 17L, degree = 3L,
  penalty_order = 3L,
  niter         = 5000L, burnin = 1000L, adapt = 500L,
  shape         = "unimodal",
  seed          = 1
)
fit_b
```

```{r, fig.cap = "Bayesian PCLM fit with unimodality constraint and 90% credible band."}
plot(fit_b)
```

### Posterior summaries

The paper (§6.1, p. 1395) reports posterior mean and 90% credible
interval for the proportion above the 30 µg/dl risk threshold, the
mean, the standard deviation, and selected quantiles. We can
reproduce them from our chain:

```{r}
prob_above_30 <- apply(fit_b$pi_chain, 1L, function(p) {
  1 - approx(x = fit_b$grid, y = c(0, cumsum(p)),
             xout = 30, rule = 2)$y
})
c(mean   = mean(prob_above_30),
  lower  = quantile(prob_above_30, 0.05, names = FALSE),
  upper  = quantile(prob_above_30, 0.95, names = FALSE))

summary(fit_b, probs = c(0.20, 0.80))
```

### Shape constraints

`bpclm()` accepts `shape = "unimodal"`, `"logconcave"`, or
`"monotonic"`, individually or as a vector. The constraint is
imposed via rejection inside the MCMC: any proposed φ whose induced
density violates the constraint is rejected outright (Eq. 7 of the
paper). For density estimation problems where the underlying
distribution is known to be unimodal — like the bloodlead example —
this stabilises the fit in the tails.

## Reproducing other paper figures

The package ships the bloodlead dataset verbatim. The
`tbdeaths1907` dataset is an *illustrative reconstruction* of the
historical CBS data used in §6.2 of Lambert and Eilers (2009); see
`?tbdeaths1907`.

```{r, fig.cap = "Bayesian fit to the (illustrative) tbdeaths1907 data."}
data(tbdeaths1907)
fit_tb <- bpclm(
  m             = tbdeaths1907$count,
  wide_breaks   = with(tbdeaths1907, cbind(lower, upper)),
  a             = 0, b = 120,
  ngrid         = 120L,
  ndx           = 17L, degree = 3L, penalty_order = 3L,
  niter         = 4000L, burnin = 1000L, adapt = 500L,
  seed          = 2
)
plot(fit_tb, xlab = "Age (years)")
```

# Part 2 — Exact ungrouping

## Why the default fit doesn't preserve bin totals

Under the Lambert–Eilers model the wide-bin counts are a multinomial
*sample* of an underlying smooth density. The smoothness penalty
shrinks the fitted bin probabilities away from the empirical
proportions, so the implied counts deviate from the observed ones.
On the bloodlead data the largest deviation is about 7.4 deaths
(roughly a 10% shrinkage of the largest bin):

```{r}
data.frame(
  lower = bloodlead$lower,
  upper = bloodlead$upper,
  m_obs = bloodlead$count,
  m_fit = round(fit_f$fitted_counts, 2L)
)
```

For a density-estimation use case this is the *correct* behaviour —
the count $m_j$ is a noisy realisation, and we are estimating the
underlying rate. For an *ungrouping* use case (where the wide-bin
counts are themselves the data — say, deaths in a 5-year age band —
and we want a fine-grid breakdown that sums back exactly), this
shrinkage is unwanted.

## Two routes to exact preservation

### `calibrate()` — post-hoc projection

Take any fit and rescale the within-bin density so $\gamma_j = m_j /
m_+$ exactly. The within-bin shape comes from the smooth fit; small
kinks may appear at bin boundaries. Bayesian draws are calibrated
one by one, so the constraint is enforced on every draw.

```{r}
fit_f_cal <- calibrate(fit_f)
data.frame(
  lower = bloodlead$lower,
  upper = bloodlead$upper,
  m_obs = bloodlead$count,
  m_fit = fit_f_cal$fitted_counts
)
```

### `pclm_exact()` — constrained MAP

Solve the constrained quadratic program
$$
\min_\phi \tfrac{1}{2}\,\phi'\,P\,\phi
\quad \text{s.t.}\quad
C\,\pi(\phi) = m / m_+,
$$
returning the smoothest density on the fine grid that exactly
reproduces the wide-bin counts. No kinks at boundaries.

```{r}
fit_e <- pclm_exact(
  m             = bloodlead$count,
  wide_breaks   = with(bloodlead, cbind(lower, upper)),
  a             = 0, b = 80,
  ngrid         = 80L, ndx = 17L, degree = 3L, penalty_order = 3L
)
max(abs(fit_e$fitted_counts - fit_e$m))     # ~ 1e-13
```

## Single-year ungrouping demonstration

Group simulated single-year mortality counts into 5-year bands and
ungroup back. This is the canonical use case in epidemiology and
demography.

```{r ungroup-setup}
ages <- 0:99
true_p <- 0.10 * dnorm(ages, mean =  2, sd =  3) +
          0.06 * dnorm(ages, mean = 25, sd = 12) +
          0.84 * dnorm(ages, mean = 75, sd = 12)
true_p <- true_p / sum(true_p)
N <- 50000L
set.seed(42)
true_yearly <- as.numeric(rmultinom(1, size = N, prob = true_p))
year_to_band <- (ages %/% 5L) + 1L
m_band <- as.numeric(tapply(true_yearly, year_to_band, sum))
wb     <- cbind(seq(0, 95, by = 5), seq(5, 100, by = 5))
```

```{r ungroup-fit}
fit_pe <- pclm_exact(m = m_band, wide_breaks = wb,
                     a = 0, b = 100, ngrid = 100L,
                     ndx = 22L, degree = 3L, penalty_order = 3L)
yhat <- fit_pe$pi * sum(m_band)
sqrt(mean((yhat - true_yearly) ^ 2))         # RMSE per single year
max(abs(fit_pe$fitted_counts - m_band))      # exact preservation
```

The RMSE per year sits close to the irreducible
multinomial-within-band sampling SD ($\sqrt{n_j / 6} \approx 20$ for
a 2500-death band). See the validation script in
`dev/validate-ungrouping.R` for a Monte Carlo coverage check.

# Part 3 — Uncertainty for ungrouping (`posterior_predict()`)

`pclm_exact()` is a deterministic point estimate — it has no
uncertainty by itself. Combining `bpclm()` with `calibrate()` gives a
chain of densities that all preserve the bin totals exactly. The
helper `posterior_predict()` then turns the chain into per-cell
credible/prediction intervals.

Two distinct uncertainty objects are supported:

* `type = "rate"` — uncertainty about the smooth latent rate
  $m_+ \pi$. Narrow at large $N$. Use this when you want to compare
  rates across populations.
* `type = "predictive"` (default) — uncertainty about the *realised*
  per-cell counts. Within each band $j$ the conditional distribution
  is multinomial of size $m_j$ with within-band probabilities
  $\pi_y / \gamma_j$. Use this when you want plausible breakdowns of
  the observed band counts.

```{r posterior-predict}
fit_b <- bpclm(m = m_band, wide_breaks = wb,
               a = 0, b = 100, ngrid = 100L,
               ndx = 22L, degree = 3L, penalty_order = 3L,
               niter = 4000L, burnin = 1000L, adapt = 500L,
               seed = 7)
fit_b <- calibrate(fit_b)               # exact band totals on every draw

pp_predictive <- posterior_predict(fit_b, type = "predictive",
                                    level = 0.9, seed = 7)
pp_rate       <- posterior_predict(fit_b, type = "rate",
                                    level = 0.9)
```

```{r posterior-predict-plot, fig.width = 8, fig.height = 4.5,
        fig.cap = "Posterior predictive (left) and rate (right) for single-year deaths."}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
plot(pp_predictive, show_bins = FALSE,
     main = "type = 'predictive' (90% PI)",
     xlab = "Age (years)", ylab = "Single-year deaths")
lines(ages, true_yearly, col = "firebrick", lwd = 1.5, lty = 2)

plot(pp_rate, show_bins = FALSE,
     main = "type = 'rate' (90% CI)",
     xlab = "Age (years)", ylab = "Expected single-year deaths")
lines(ages, true_yearly, col = "firebrick", lwd = 1.5, lty = 2)
par(op)
```

The predictive interval (left) is wider because it includes the
multinomial sampling noise; it has approximately nominal coverage
(≈ 85% empirical at the 90% nominal level in our validation). The
rate interval (right) is much narrower because at $N = 50000$ the
smooth density is well-determined and the realised counts have
substantial sampling noise.

## Cumulative sums

Because the fine-cell draws preserve every band total exactly, any
cumulative sum at a band boundary is exactly known: the credible
band closes onto the truth at every multiple of 5. *Within* a band
the credible band has positive width, reflecting how the band's
counts could have been distributed across single years.

```{r cumulative-fig, fig.cap = "Cumulative-deaths posterior predictive."}
cdf_chain <- t(apply(pp_predictive$draws, 1L, cumsum))
cdf_mean  <- colMeans(cdf_chain)
cdf_lo    <- apply(cdf_chain, 2L, quantile, probs = 0.05, names = FALSE)
cdf_hi    <- apply(cdf_chain, 2L, quantile, probs = 0.95, names = FALSE)

plot(ages, cumsum(true_yearly), type = "l", col = "grey40", lwd = 2,
     xlab = "Age (years)", ylab = "Cumulative deaths",
     main = "Cumulative deaths: truth + 90% predictive band")
polygon(c(ages, rev(ages)), c(cdf_lo, rev(cdf_hi)),
        col = grDevices::adjustcolor("steelblue", 0.3), border = NA)
lines(ages, cdf_mean, col = "black", lwd = 2)
abline(v = seq(5, 95, by = 5), col = "grey90", lty = 3)
legend("topleft",
       legend = c("Truth", "Posterior predictive mean",
                   "90% PI", "5-year band boundaries"),
       col = c("grey40", "black",
                grDevices::adjustcolor("steelblue", 0.6), "grey90"),
       lwd = c(2, 2, 8, 1), lty = c(1, 1, 1, 3), bty = "n")
```

## When to use which

| Goal | Function |
|---|---|
| Smooth density from sampled data, full Bayesian inference | `bpclm()` |
| Same, but with shape constraints (unimodal, log-concave, monotonic) | `bpclm(..., shape = ...)` |
| Smooth density from sampled data, fast point estimate | `pclm()` |
| Ungrouping: smoothest density that *exactly* preserves bin totals (point estimate) | `pclm_exact()` |
| Ungrouping with uncertainty: posterior chain that exactly preserves bin totals | `bpclm()` |&#8203;`%>%`&#8203;`calibrate()` |
| Per-cell credible / prediction intervals for ungrouped counts | `posterior_predict(fit, type = "predictive")` |
| Per-cell credible interval for the smooth rate only | `posterior_predict(fit, type = "rate")` |

# References

Lambert, P. and Eilers, P. H. C. (2009). Bayesian density estimation
from grouped continuous data. *Computational Statistics and Data
Analysis*, 53(4), 1388–1399. doi:10.1016/j.csda.2008.11.022

Eilers, P. H. C. and Marx, B. D. (1996). Flexible smoothing with
B-splines and penalties. *Statistical Science*, 11(2), 89–121.

Hasselblad, V., Stead, A. G. and Galke, W. (1980). Analysis of
coarsely grouped data from the lognormal distribution. *Journal of the
American Statistical Association*, 75, 771–778.

Roberts, G. O. and Rosenthal, J. S. (1998). Optimal scaling of
discrete approximations to Langevin diffusions. *Journal of the Royal
Statistical Society, Series B*, 60(1), 255–268.

Atchadé, Y. F. and Rosenthal, J. S. (2005). On adaptive Markov chain
Monte Carlo algorithms. *Bernoulli*, 11(5), 815–828.

Camarda, C. G., Eilers, P. H. C. and Pascariu, M. (2018). ungroup:
An R package for efficient estimation of smooth distributions from
coarsely binned data. *Journal of Open Source Software*, 3(29), 937.
